Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2023-05-01 analysis/paper/paper.qmd
Edited based on Lape's comments

Since stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origins is critical to understanding the technological transition from the Early Paleolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern]. The debate around the origin of stemmed points can be summarized into two competing models [@bae2013early;@lee2016patterns] : (1) a 'heterogenic' migration [@bae2010origin], and (2) an *in situ* evolution model [@Seong_2006]. The migration model claims that the new blade industry - including stemmed points - and the earlier coarse flake tradition - including large cores, polyhedrons, choppers, and even handaxes - came from different origins. These are (1) a Northern route: Siberia, Mongolia, or other regions of northeast China, and (2) a Southern route: southern China) as the result of a continuing influx of modern human migration from two routes. The *in situ* model argues that stemmed points and other Late Paleolithic technologies, such as blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence. The difference between the two models comes from how they explain a few early sites and stemmed points made out of flakes. The *in situ* model argues that stemmed points appeared before blades, and identifies early sites such as Bonggok, Songamri, Yonghodong, and Hwadaeri as evidence that the Korean Late Paleolithic began with the emergence of stemmed points. In contrast, the migration model contends that the Late Paleolithic began with the introduction and widespread use of blade technology, similar to the traditional definition of the Late or Upper Paleolithic in Europe and the Later Stone Age in Africa. [@seong2016eastern]. Besides these two main models, there are combination of migration and trade interaction model [@bae2012nature] and complex and non-directional model [@lee2016patterns].

@@

To answer our research questions about the likely uses for stemmed points, we calculated TCSA values for the stemmed points in our sample. We also explored the interaction of TCSA values with raw materials and artifact size, using weight as a proxy. Because the shape of stone artifacts is highly influenced by the raw materials [@mcpherron2014experimental], we assumed that raw materials may be highly correlated with tool size. We investigated radiocarbon ages associated with the points in order to determine temporal patterns in TCSA values. We separated the research time period into three phases based on the major climate event during the Late Pleistocene, the Last Glacial Maximum (LGM), and examined the distribution of TCSA values before, during, and after the LGM. We explored the relationship between TCSA values and the location of assemblages by comparing the distribution of TCSA values across ecological and vegetation zones.

@@

```{r}
#| label: fig-tcsa-all-sp
#| fig-cap: "Distribution of TCSA values for all Korean stemmed points in the current dataset. Shaded boxes indicate TCSA ranges for various weapon types."
#| fig-width:   3
#| fig-height:  2

@@

ggplot(tcsa,
        aes(0.3, 
            min_tcsa)) +
  # show regions for tools from Lomabrd 2021 & 2022
  # https://www-sciencedirect-com.offcampus.lib.washington.edu/science/article/pii/S0277379121002286
  # https://assets.researchsquare.com/files/rs-2016964/v1/00d381a3-3ecf-4d5b-b2ec-89a41227bbd5.pdf?c=1662475267
    annotate("rect", 
           xmin = 0.1,
           xmax = 0.2,
           ymin = 4,
           ymax = 18,
           fill = "green",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Poisoned arrow tips",
               hjust = 0,
               lineheight = 0.75,
           x = point_function_group_label_y,
           y = (18-4) /2 + 4) +
      annotate("rect", 
           xmin = 0.2,
           xmax = 0.3,
           ymin = 17,
           ymax = 47,
           fill = "blue",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Un-poisoned arrow tips",
               hjust = 0,
               lineheight = 0.75,
           x = point_function_group_label_y,
           y = (47-17) /2 + 17) +
      annotate("rect", 
           xmin = 0.3,
           xmax = 0.4,
           ymin = 40,
           ymax = 76,
           fill = "orange",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Dart tips",
               hjust = 0,
           x = point_function_group_label_y,
           y = (76-40) /2 + 40 - 5) +
      annotate("rect", 
           xmin = 0.4,
           xmax = 0.5,
           ymin = 42,
           ymax = 90,
           fill = "purple",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Javelin tips",
               hjust = 0,
           x = point_function_group_label_y,
           y = (90-42) /2 + 42 ) +
     annotate("rect", 
           xmin = 0.5,
           xmax = 0.6,
           ymin = 80,
           ymax = 200,
           fill = "yellow",
           alpha = 0.2) +
      annotate("text", 
               label = "Stabbing spear tips",
               size = point_function_group_label_size,
               hjust = 0,
               lineheight = 0.75,
           x = point_function_group_label_y,
           y = (200-80) /2 + 80 ) +
  geom_boxplot(alpha = 0.1) +
  geom_quasirandom(alpha = 0.4,
                   size = 0.7) +
  coord_cartesian(xlim = c(-0.1,1)) +
  theme_bw(base_size = 6) +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()
        ) +
  labs(y = "TCSA")

@@

############ explore tcsa with MODERN eco_regions #######
library(sf)
veg_all <- read_sf(here("analysis/data/gis_data/eco_re_two.shp"))
veg_all <- st_transform(veg_all, crs = st_crs(4326))
# ggplot() +
#   geom_sf(data = veg_all) +
#   coord_sf()
pts_all <- st_as_sf(korean_archaeological_site_locations_tcsa,
                coords = c("long_dd.y", "lat_dd.y"),
                crs = st_crs(veg_all))
# ggplot() +
#   geom_sf(data = pts_all)
sf_use_s2(FALSE) # to prevent error. Check here for more info: https://github.com/r-spatial/sf/issues/1762
points_in_veg_zones_all <- 
st_join(veg_all, 
        pts_all, 
        join = st_contains)
eco_plot <- ggplot() +
  geom_sf(data = veg_all, aes(fill = "grey")) +
   geom_sf_text(data = veg_all, aes(label = ecoregion), size =1.5)+
  geom_sf(data = pts_all, aes(color = min_tcsa), size = 1.1) +
  theme_minimal() +
  scale_fill_brewer(palette="Greys") +
  labs(color="TCSA") +
  guides(fill = "none") +
  theme(axis.text.x = element_text(angle = 45, size = 4.5))+
  theme(axis.text.y = element_text(size = 4.5))+
  theme(axis.title.x = element_blank(),
          axis.title.y = element_blank()) +
  theme(legend.position = "none")

@@

# ggplot() +
#   geom_sf(data = veg_forest) +
#   coord_sf()
pts_forest <- 
  st_as_sf(korean_archaeological_site_locations_tcsa,
                coords = c("long_dd.y", "lat_dd.y"),
                crs = st_crs(veg_forest))
# ggplot() +
#   geom_sf(data = pts_forest)
#sf_use_s2(FALSE)
points_in_veg_zones_forest <- 
st_join(veg_forest, 
        pts_forest, 
        join = st_contains)
for_plot <- ggplot() +
  geom_sf(data = veg_forest, aes(fill = "grey")) +
  geom_sf_text(data = veg_forest, aes(label = Zone), size =1.5)+
  geom_sf(data = pts_forest, aes(color = min_tcsa), size = 1.1) +
  theme_minimal()+
  scale_fill_brewer(palette="Greys") +
  labs(color="TCSA") +
  guides(fill = "none") +
  theme(axis.text.x = element_text(angle = 45, size = 4.5)) +
  theme(axis.text.y = element_text(size = 4.5)) +
  theme(axis.title.x = element_blank(),
          axis.title.y = element_blank()) +
  theme(legend.title = element_text(size= 6),
        legend.key.height = unit(4.5, 'mm'),
        legend.key.width = unit(2, 'mm'),
        legend.text = element_text(size = 4))

@@

By comparing the results with the TCSA range from other archaeological and ethnographic cases (@tbl-tcsa-ranges), we were able to determine what role stemmed points played in the technological transition during the Korean Late Paleolithic. Our main questions were: What were the best-fit ballistic probabilities for stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns of stemmed point uses?

@@

According to the different weapon-delivery systems that can be inferred from TCSA values, the majority of stemmed points from the Korean Paleolithic were probably used as javelin tips and stabbing spear tips. In general, though, we noted a wide range and also differing distributions of TCSA values in each assemblage. Therefore, we conclude that stemmed points served diverse and not highly patterned roles during the Late Paleolithic. Prior to the LGM, people may have encountered unfamiliar situations and use stemmed points as a multifunctional tool to carry out multiple tasks. During the LGM and following we find production and discard of stemmed points declined, but their ballistic properties remained largely unchanged. Considering that many of our predictions about TCSA were not supported by the data, we speculate that composite projectile points made with micro-blades were newly introduced in response to global climate dynamics and stemmed points were still optimal for the wide range of conditions encountered by people during the Late Paleolithic in Korea.

We are aware that discriminating the likely use of small numbers of projectile points could be arbitrary [@erlandson2014darts]. Since TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e. increase or decrease by shape of tip and cross section), we nevertheless consider it a useful metric for hypothesizing about different weapon-delivery systems [@hughes1998getting; @sitton2020tip; @lombard2021variation]. As part of our future research, we plan to conduct use-wear analyses of stemmed points and experiments in an effort to elucidate in more detail the function of stemmed points and technological links between hunter-gatherer groups.

>>>>

Since the first appearance of stemmed points defines the beginning of the Korean Late Paleolithic, investigating their origins is critical to understanding the technological transition from the Early Paleolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern]. The debate around the origin of stemmed points can be summarized into two competing models [@bae2013early;@lee2016patterns] : (1) a 'heterogenic' migration [@bae2010origin], and (2) an *in situ* evolution model [@Seong_2006]. The migration model claims that the new blade industry - including stemmed points - and the earlier coarse flake tradition - including large cores, polyhedrons, choppers, and even handaxes - came from different origins. These are (1) a Northern route: Siberia, Mongolia, or other regions of northeast China, and (2) a Southern route: southern China) as the result of a continuing influx of modern human migration from two routes. The *in situ* model argues that stemmed points and other Late Paleolithic technologies, such as blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence. The difference between the two models comes from how they explain a few early sites and stemmed points made out of flakes. The *in situ* model argues that stemmed points appeared before blades, and identifies early sites such as Bonggok, Songamri, Yonghodong, and Hwadaeri as evidence that the Korean Late Paleolithic began with the emergence of stemmed points. In contrast, the migration model contends that the Late Paleolithic began with the introduction and widespread use of blade technology, similar to the traditional definition of the Late or Upper Paleolithic in Europe and the Later Stone Age in Africa. [@seong2016eastern]. Besides these two main models, there are combination of migration and trade interaction model [@bae2012nature] and complex and non-directional model [@lee2016patterns].

@@

To answer our research questions about the likely uses for stemmed points, we calculated TCSA values for the stemmed points in our sample. We also explored the interaction of TCSA values with raw materials and artifact size, using weight as a proxy. Because the shape of stone artifacts is highly influenced by the raw materials [@mcpherron2014experimental], we assumed that raw materials may be highly correlated with tool size. We investigated radiocarbon ages associated with the points in order to determine temporal patterns in TCSA values. We separated the research time period into three phases, based on the major climate event during the Late Pleistocene, the Last Glacial Maximum (LGM). Previous research on the LGM climate in Korea, using age-controlled pollen records [@yi2010vegetation] and computational simulation models [@kim2015last;@park2022did], indicates that the climate was colder and drier than the preceding period. Surface temperature cooling ranged from 5 to 6°C, and there was a precipitation decrease of approximately 14%. Using this chronology, we examined the distribution of TCSA values before, during, and after the LGM. We then explored the relationship between TCSA values and the location of assemblages by comparing the distribution of TCSA values across ecological and vegetation zones.

@@

```{r}
#| label: fig-tcsa-all-sp
#| fig-cap: "Distribution of TCSA values for all Korean stemmed points in the current dataset. The shaded boxes in color indicate TCSA ranges for different weapon types based on @tbl-tcsa-ranges."
#| fig-width:   3
#| fig-height:  2

@@

ggplot(tcsa,
        aes(0.3, 
            min_tcsa)) +
  # show regions for tools from Lomabrd et al. 2022
  # https://www-sciencedirect-com.offcampus.lib.washington.edu/science/article/pii/S0277379121002286
  # https://assets.researchsquare.com/files/rs-2016964/v1/00d381a3-3ecf-4d5b-b2ec-89a41227bbd5.pdf?c=1662475267
    annotate("rect", 
           xmin = 0.1,
           xmax = 0.2,
           ymin = 4,
           ymax = 18,
           fill = "green",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Poisoned arrow tips",
               hjust = 0,
               lineheight = 0.75,
           x = point_function_group_label_y,
           y = (18-4) /2 + 4) +
      annotate("rect", 
           xmin = 0.2,
           xmax = 0.3,
           ymin = 17,
           ymax = 47,
           fill = "blue",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Un-poisoned arrow tips",
               hjust = 0,
               lineheight = 0.75,
           x = point_function_group_label_y,
           y = (47-17) /2 + 17) +
      annotate("rect", 
           xmin = 0.3,
           xmax = 0.4,
           ymin = 40,
           ymax = 76,
           fill = "orange",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Dart tips",
               hjust = 0,
           x = point_function_group_label_y,
           y = (76-40) /2 + 40 - 5) +
      annotate("rect", 
           xmin = 0.4,
           xmax = 0.5,
           ymin = 42,
           ymax = 90,
           fill = "purple",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Javelin tips",
               hjust = 0,
           x = point_function_group_label_y,
           y = (90-42) /2 + 42 ) +
     annotate("rect", 
           xmin = 0.5,
           xmax = 0.6,
           ymin = 80,
           ymax = 200,
           fill = "yellow",
           alpha = 0.2) +
      annotate("text", 
               label = "Stabbing spear tips",
               size = point_function_group_label_size,
               hjust = 0,
               lineheight = 0.75,
           x = point_function_group_label_y,
           y = (200-80) /2 + 80 ) +
  geom_boxplot(alpha = 0.1) +
  geom_quasirandom(alpha = 0.4,
                   size = 0.7) +
  coord_cartesian(xlim = c(-0.1,1)) +
  theme_bw(base_size = 6) +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()
        ) +
  labs(y = "TCSA")

@@

############ explore tcsa with MODERN eco_regions #######
library(sf)
veg_all <- read_sf(here("analysis/data/gis_data/eco_re_two.shp"))
veg_all <- st_transform(veg_all, crs = st_crs(4326))
# ggplot() +
#   geom_sf(data = veg_all) +
#   coord_sf()
pts_all <- st_as_sf(korean_archaeological_site_locations_tcsa,
                coords = c("long_dd.y", "lat_dd.y"),
                crs = st_crs(veg_all))
# ggplot() +
#   geom_sf(data = pts_all)
sf_use_s2(FALSE) # to prevent error. Check here for more info: https://github.com/r-spatial/sf/issues/1762
points_in_veg_zones_all <- 
st_join(veg_all, 
        pts_all, 
        join = st_contains)

eco_plot <- ggplot() +
  geom_sf(data = veg_all, aes(fill = "grey")) +
   geom_sf_text(data = veg_all, aes(label = ecoregion), size =1.2)+
  geom_sf(data = pts_all, aes(color = min_tcsa), size = 0.8) +
  theme_minimal() +
  scale_fill_brewer(palette="Greys") +
  labs(color="TCSA") +
  guides(fill = "none") +
  theme(axis.text.x = element_text(angle = 45, size = 4.5))+
  theme(axis.text.y = element_text(size = 4.5))+
  theme(axis.title.x = element_blank(),
          axis.title.y = element_blank()) +
  theme(legend.position = "none")

@@

# ggplot() +
#   geom_sf(data = veg_forest) +
#   coord_sf()
pts_forest <- 
  st_as_sf(korean_archaeological_site_locations_tcsa,
                coords = c("long_dd.y", "lat_dd.y"),
                crs = st_crs(veg_forest))
# ggplot() +
#   geom_sf(data = pts_forest)
#sf_use_s2(FALSE)
points_in_veg_zones_forest <- 
st_join(veg_forest, 
        pts_forest, 
        join = st_contains)
for_plot <- ggplot() +
  geom_sf(data = veg_forest, aes(fill = "grey")) +
  geom_sf_text(data = veg_forest, aes(label = Zone), size =1.2)+
  geom_sf(data = pts_forest, aes(color = min_tcsa), size = 0.8) +
  theme_minimal()+
  scale_fill_brewer(palette="Greys") +
  labs(color="TCSA") +
  guides(fill = "none") +
  theme(axis.text.x = element_text(angle = 45, size = 4.5)) +
  theme(axis.text.y = element_text(size = 4.5)) +
  theme(axis.title.x = element_blank(),
          axis.title.y = element_blank()) +
  theme(legend.title = element_text(size= 6),
        legend.key.height = unit(4.5, 'mm'),
        legend.key.width = unit(2, 'mm'),
        legend.text = element_text(size = 4))

@@

By comparing the results with the TCSA ranges from other archaeological and ethnographic cases (@tbl-tcsa-ranges), we were able to examine what role stemmed points played in the technological transition during the Korean Late Paleolithic. Our main questions were: What were the best-fit ballistic probabilities for stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns of stemmed point uses?

@@

According to the different weapon-delivery systems that can be inferred from TCSA values, the majority of stemmed points from the Korean Paleolithic were probably used as javelin tips and stabbing spear tips. In general, though, we noted a wide range and also differing distributions of TCSA values in each assemblage. Therefore, we conclude that stemmed points served diverse and not highly patterned roles during the Late Paleolithic. Prior to the LGM, people may have encountered unfamiliar situations and use stemmed points as a multifunctional tool to carry out multiple tasks. During the LGM and following we find production and discard of stemmed points declined, but their ballistic properties remained largely unchanged. Considering that many of our predictions about TCSA were not supported by the data, we speculate that composite projectile points made with microblades were newly introduced [@chang2013human], in response to global climate dynamics and stemmed points were still optimal for the wide range of conditions encountered by people during the Late Paleolithic in Korea.

We are aware that discriminating the likely use of small numbers of projectile points could be arbitrary [@erlandson2014darts]. Since TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e. increase or decrease by shape of tip and cross section), we nevertheless consider it a useful metric for hypothesizing about different weapon-delivery systems [@hughes1998getting; @sitton2020tip; @lombard2021variation]. As part of our future research, we plan to conduct use-wear analyses and experiments in an effort to elucidate in more detail the function of stemmed points and technological links between hunter-gatherer groups.
